Network Origins of Social Distinction

### 2025-07-30 manuscript.tex

In addition to these explanations, the distinction measure also reveals the utility of a distinction based theory of triadic closure. Figure 3 \textit{Model (a)} and Table 4 present a graph and corresponding measures of near-neighbor triadic closure. The distinction of nodes 1 and 6 increases to .106, compared to 0 in the circle graph, while all other nodes fall below 0 due to the reverberating effects of the new connection. Notably, node 7 increases in eigenvector status from the combination, but overall decrease in scaled distinction due to the tie. While both eigenvector centrality and scaled distinction would predict tie formation in this simplified social structure, the scaled distinction measure would find that Node 7 would actually lose out in some ways from this arrangement. This matches with findings that initial connections do weaken when triadic closure occurs \citep{huang2018will}, consistent with a reaction to the increased constraint placed on the closure broker. This example demonstrates the fragility of the circle graph as an equality condition in a distinction-based theoretical framework, as there are opportunities for mutual increases in distinction through triadic closure. Additionally, the  example of triadic closure shows that only mild additions to network ties can very quickly create stratification. The addition of a single connection between node 1 and 6 in Figure 3 \textit{Model (a)} stratified a network from equality into four levels of hierarchy with a single additional connection.

Beyond questions of balancing valence as is the current framework for understanding triadic closure, a distinction-based theoretical framework can explain the propensity towards triadic closure, but also its failure to form depending on the initial patterning of relationships.

@@

What the difference between these two examples suggests is that, based on considerations for distinction, the initial spread of social organization occurs locally through the triadic balance of near neighbors. This would suggest that the natural form of organizational formation is the pyramid formed from combinations of local triads that are further connected by brokers or welds creating structural folds. However, the transitivity created by the initial triad formation would spread quickly throughout the network, making welding structural folds a highly advantageous strategy for members of triads who could facilitate that connection, which is further explored below.

>>>>

In addition to these explanations, the distinction measure also reveals the utility of a distinction based theory of triadic closure. Figure 3 \textit{Model (a)} and Table 4 present a graph and corresponding measures of near-neighbor triadic closure. The distinction of nodes 1 and 6 increases to .106, compared to 0 in the circle graph, while all other nodes fall below 0 due to the reverberating effects of the new connection. Notably, node 7 increases in eigenvector status from the combination, but overall decrease in scaled distinction due to the tie. This example demonstrates the fragility of the circle graph as an equality condition in a distinction-based theoretical framework, as there are opportunities for mutual increases in distinction through triadic closure. While both eigenvector centrality and scaled distinction would predict tie formation in this simplified social structure, the scaled distinction measure would find that Node 7 would actually lose out in some ways from this arrangement. This matches with findings that often initial connections do weaken when triadic closure occurs \citep{huang2018will}, consistent with a reaction to the increased constraint placed on the closure broker. However, the actual incentives and consequences for each of the nodes in a triadic closure will be dependent on the overall existing network structure. Node 7 is worse off because of the structure of the circle graph, not necessarily because in all conditions Node 7 will be worse off. If Node 7 is actually highly central to the graph, near neighbor triadic closure can be distinction increasing. Beyond questions of balancing valence as is the current framework for understanding triadic closure, a distinction-based theoretical framework can explain the propensity towards triadic closure, but also its failure to form depending on the initial patterning of relationships.

@@

What the difference between these two examples suggests is that, based on considerations for distinction, the initial spread of social organization occurs locally through the triadic balance of near neighbors. This would suggest that the natural form of organizational formation is the pyramid formed from combinations of local triads that are further connected by brokers or welds creating structural folds. However, the transitivity created by the initial triad formation would spread quickly throughout the network, making welding structural folds a highly advantageous strategy for members of triads who could facilitate that connection, which is further explored below. Additionally, the example of triadic closure shows that only mild additions to network ties can very quickly create stratification. The addition of a single connection between node 1 and 6 in Figure 3 \textit{Model (a)} stratified a network from equality into four levels of hierarchy with a single additional connection.
